---
title: "`data.frame` vs `tibble`"
date: "2025-04-15"
categories: [R, base::data.frame, tibble::tibble, tibble::tribble]
---

# Create

Base R provides a data structure, called `data.frame`, to store and manipulate
tabular data.

A `data.frame` is a `list` (a R list!) of equal-length vectors.

Creating a `data.frame` is as easy as

```{r}
df <-
  base::data.frame(
    letters = letters,
    numbers = 1:26
  )
```

A `tibble` is the tidyverse version of `data.frame`.

Using similar syntax we can create a `tibble`:

```{r}
tbl <-
  tibble::tibble(
    letters = letters,
    numbers = 1:26
  )
```

# Differences

## Print

The first difference between `data.frame` and `tibble` is about printing:
tibbles have a print method that shows only the first 10 rows, and all the
columns that fit on screen. This makes it much easier to work with large data.
In addition to its name, each column reports its type.

```{r}
df
tbl
```


## Subset and Partial Matching

The second difference is about subsetting. Data frames allow partial matching
for column names:

```{r}
df$let
```

On the other hand a tibble is stricter and does not allow partial matching:

```{r}
tbl$let
```

This can be very helpful, especially when debugging code, because the message
tells use where the problem is. Look at what happens in the following example:

```{r}
df2 <- base::data.frame(
  letters_lower = letters,
  letters_upper = LETTERS,
  numbers = 1:26
)

df2$let
```

Since more than one column name partially matches the string 'let', subsetting
returns `NULL` without generating a warning. It would be more difficult to debug
a bigger chunk of code if this technique is used. Therefore it’s advisable to
avoid partial matching when subsetting. Tibbles are good because they forbid us
to use it.

## Extract a single column with `[`

Subetting a single element from a `data.frame` using `[` does not return a
`data.frame`.

```{r}
df_subset <- df[, "letters"]

base::is.data.frame(df_subset)

base::class(df_subset)

utils::str(df_subset)
```

Conversly, subetting a single element from a `tibble` using `[` returns a
`tibble`.

```{r}
tbl_subset <- tbl[, "letters"]

tibble::is_tibble(tbl_subset)

base::class(tbl_subset)

utils::str(tbl_subset)
```

## Add a New Column Using `$`

Adding a new column to a `data.frame` using `$` will recycle the vector used if
its length is less than the number of rows of the `data.frame` (but assuming
that the number of elemnts of the vector is a submultiple of the number of rows
of the `data.frame`).

```{r}
df$other_numbers <- 27:28

df$other_numbers

df
```

In general, a `tibble` does not allow it

```{r}
#| error: true
tbl$other_numbers <- 27:28
```

Unless we want to add a constant column

```{r}
tbl$constant_column <- 7

tbl$constant_column

tbl
```

Note that if the length of the vector used to define a new column is equal to
the number of rows of the tibble, then everything will run smoothly.

```{r}
#| echo: false
tbl <-
  tibble::tibble(
    letters = letters,
    numbers = 1:26
  )
```

```{r}
tbl$other_numbers <- 27:52

tbl$other_numbers

tbl
```

# Create by Row

A useful function coming from the `tibble` package is `tribble` which allow us
to define a `tibble` by row. This is useful for small tables of data where
readability is important.

```{r}
tibble::tribble(
  ~letters, ~numbers,
  "a",      1,
  "b",      2,
  "c",      3
)
```

There are some workarounds to create a `data.frame` by row, however they do not

# References

- [Tibble](https://tibble.tidyverse.org/index.html) package documentation

- [Tibble](https://posit.co/blog/tibble-1-0-0/) from Posit blog
